Avoiding tipping points in the management of ecological systems: a non-parametric Bayesian approach

### 2013-06-24 inst/doc/nonparametric-bayes.Rmd
writing

<!-- Models for decision-making -->
Decision-theoretic tools require a model that can assign probabilites of
future states (e.g. stock size of a fishery) given the current state and a
proposed action (e.g. fishing harvest or effort).  The decision maker then
seeks to determing the course of actions (also refered to as the policy)
that maximizes the expected value of some objective function (such as
net present value derived from the resource over time.  (The approach
can be adapted to permit alternatives to maximizing expectation, such
as minimizing the maximum cost or damage that might be incurred. See
@Polasky2011).
Management frequently faces a sequential decision-making problem -- after 
selecting an action, the decision-maker may recieve new information about 
the current state and must again choose an appropriate action -- i.e. setting
the harvest limits each year based on stock assessments the year prior.

@@

<!-- Parametric uncertainty -->

<!-- Comparing models -->
The nature of decision-making problems provides a convenient way to compare 
models.  Rather than compare models in terms of best fit to data or fret over
the appropriate penalty for model complexity, model performance is defined 
in the concrete terms of the decision-maker's objective function, which we
will take as given. (Much argument can be made over the 'correct' objective
function, e.g. how to account for the social value of fish left in the sea
vs. the commercial value of fish harvested; see @Halpern2013 for further 
discussion of this issue.  Alternatively, we can always compare model performance
across multiple potential objective functions.)  The decision-maker
does not necessarily need a model that provides the best mechanistic understanding
or the best long-term outcome, but rather the one that best estimates the 
probabilites of being in different states as a result of the possible actions.

<!-- Structural Uncertainty -->

@@

### Quantitative vs Qualitative Decisions

@@

Non-parametric models can more realistically represent uncertainties outside 
of observed data while also better capturing the dynamics in the region
observed.  Advances in the theory and computational implementations of 
nonparametric methods make them ripe for such applications.  We use the classic
problem of optimal harvest of a marine fishery to illustrate how the 
nonparametric approach of Gaussian processes can be applied.  We will 
compare Bayesian implementations of both nonparametric and parametric 
models, which best allow us to capture the uncertainty of model estimation
in either case and permits a more natural comparison between approaches.

@@

### Uncertainties: the classical approach

The ecological dynamics of most systems of management interest are
typically both complex and unknown.  Despite this, quantitative
policy optimization in these systems has almost exclusively been based
on simple, mechanistically motivated parametric models.  There are
several reasons for this.  First, limited available data almost always
precludes any accurate estimate of more complicated models.  Second, the
computational demands of determining an optimal strategy in a sequential
decision-theory problem suffer from the "curse of dimensionality"
that makes it difficult to consider more complicated models, states,
or actions.  Unfortunately, these challenges also make it particularly
difficult to reflect the true uncertainty either in the parameters of
the given models (parametric uncertainty) or in the structure of the
models themselves (structural uncertainty).  Further unknowns such as
measurement uncertainty, parameter uncertainty, unobserved states,
knowledge of boundary conditions, etc. further compound the issue.
Though a hierarchical Bayesian approach provides a natural way to
address these from a statistical standpoint, formulating reasonable
parametric  descriptions of each form of uncertainty is a challenging
task in itself, let alone the computational difficulties of solving such
a system. @Cressie2009 provides a good account of the successes and
challenges of the approach.  Applying these approaches in the management
context of sequential decision making, in which forecasts must be obtained
over a range of possible actions and updated regularly as new information
arrives makes such an approach less feasible still.

Traditional approaches to optimal control (Pontryagin's principle, stochastic
dynamic programming) rely on knowledge of the state equation, usually described
by a simple parametric model. Here we illustrate how a stochastic dynamic
programming algorithm can alternatively be driven by the predictions from
a Gaussian process.

@@

To illustrate the value of the non-parametric Bayesian approach to management,
we focus on example of a system containing such a tipping point whose dynamics
can still be described by a simple, one-dimensional parametric model.  
We will focus on a simple parametric model for a single species [derived
from fist principles by @Allen2005a] as our underlying "reality".

\begin{align}
X_{t+1} &= Z_t f(S_t) \\
S_t &= X_t - h_t \\
f(S_t) &= S_t e^{r \left(1 - \frac{S_t}{K}\right)\left(S_t - C\right)}
\end{align}

@@

In our example, we focus on the problem in which a manager must set 
the harvest level for a marine fishery each year to maximize the net
present value of the resource, given an estimated stock size from the year
before. Rich data and global concerns have made marine fisheries the crucible for much
of the founding work [@Gordon1954; @Reed1979; @May1979; @Ludwig1982]
in managing ecosystems under uncertainty.  Global trends [@Worm2006]
and controversy [@Hilborn2007; @Worm2009] have made understanding these
challenges all the more pressing.

@@

A crux of this approach is correctly specifying the functional form of $f$,
along with its parameters.  The standard approach uses one of a 
handful of common parametric models representing the stock-recruitment
relationship, usually after estimating the model parameters from any 
available existing data. Uncertainty in the parameter estimates can 
be estimated and integrated over to determine the optimal policy under
under uncertainty [@Mangel1985; @Schapaugh2013]. Uncertainty in the model
structure itself can only be addressed in this approach by hypothesizing 
alternative model structures, and then performing some model choice or
model averaging  [@Williams2001; @Athanassoglou2012]. For simplicity 
of the comparison we consider only the case of a structurally correct
and structurally incorrect model, and estimate Bayesian posterior distributions
of the parameters.  We will also compare to a simpler maximum likelihood
estimate of the parameters that ignores the issue of parameter uncertainty.

We compare these parametric approaches to our alternative that uses a Gaussian Process (GP)
in place of a given stock recruitment curve.  We briefly revisit
a description of the Gaussian Process approach that has appeared in 
recent literature before describing how we take the important step of 
adapting the GP to the decision theory problem that can still be solved
using SDP.

@@

The use of Gaussian process (GP) regression (or "kreging" in the geospatial
literature) to formulate a predictive model is relatively new in the
context of modeling dynamical systems [@Kocijan2005] and introduced
in the ecological modeling and fisheries management by @Munch2005.
An accessible and thorough introduction to the formulation and use of
GPs can be found in @Rasmussen2006.

>>>>

Here we illustrate how a stochastic dynamic programming algorithm can 
be driven by the predictions from a Gaussian process model, sidestepping
the need for an accurate model-based description of the sustem dynamics.

<!-- Models for decision-making -->
Decision-theoretic (or optimal control) tools require a model that can assign probabilities of
future states (e.g. stock size of a fishery) given the current state and a
proposed action (e.g. fishing harvest or effort).  The decision maker then
seeks to determining the course of actions (also referred to as the policy)
that maximizes the expected value of some objective function (such as
net present value derived from the resource over time.  (The approach
can be adapted to permit alternatives to maximizing expectation, such
as minimizing the maximum cost or damage that might be incurred. See
@Polasky2011).
Management frequently faces a sequential decision-making problem -- after 
selecting an action, the decision-maker may receive new information about 
the current state and must again choose an appropriate action -- i.e. setting
the harvest limits each year based on stock assessments the year prior.

@@

Traditional approaches to optimal control (Pontryagin's principle, stochastic
dynamic programming) rely on knowledge of the state equation.  Such simple 
parametric models, frequently justified by mechanistic underpinnings are used
to provide probabilities of future states.  Fisheries literature, which enjoys
some of the longest history in the use of the optimal control framework, provides
a clear example. Models such as the Ricker or Beverton Holt curves are ubiquitous [too strong? references?]
Parametric models can reflect variability introduced by a stochastic environment
or demographic process my incorporating random shocks from a given distribution.

<!-- Parametric uncertainty -->
As the parameter values for these models must be estimated from limited data,  
there will always be some uncertainty associated with these values.  This uncertainty
further compounds the intrinsic variability introduced by demographic or environmental 
noise.  The degree of uncertainty in the parameter values can be inferred from the data
and reflected in the estimates of the transition probabilities.

<!-- Structural Uncertainty --> 
Unfortunately, estimates of parameter uncertainty are only as good as the parametric 
models themselves.  Too often, we do not understand the system dynamics well enough
to know if a model provides a good approximation over the relevant range of states
and timescales (criteria that we loosely refer to as defining the "right" or "true" model.)
So called structural or model uncertainty is a more difficult problem than parametric 
uncertainty.  Typical solutions involve either model choice, or model averaging. Either
approach remains limited by the degree to which any of the proposed models are sufficiently
close to the right model.

<!-- Comparing models -->
The nature of decision-making problems provides a convenient way to compare 
models.  Rather than compare models in terms of best fit to data or fret over
the appropriate penalty for model complexity, model performance is defined 
in the concrete terms of the decision-maker's objective function, which we
will take as given. (Much argument can be made over the 'correct' objective
function, e.g. how to account for the social value of fish left in the sea
vs. the commercial value of fish harvested; see @Halpern2013 for further 
discussion of this issue.  Alternatively, we can always compare model performance
across multiple potential objective functions.)  The decision-maker
does not necessarily need a model that provides the best mechanistic understanding
or the best long-term outcome, but rather the one that best estimates the 
probabilities of being in different states as a result of the possible actions.

<!-- The weakness of parametric models -->

@@

<!--
### Quantitative vs Qualitative Decisions

@@

-->

Non-parametric models can better represent uncertainties outside 
of observed data while also better capturing the dynamics in the region
observed.  Advances in the theory and computational implementations of 
nonparametric methods make them ripe for such applications.  We use the classic
problem of optimal harvest of a marine fishery to illustrate how the 
nonparametric approach of Gaussian processes can be applied.  We will 
compare Bayesian implementations of both nonparametric and parametric 
models, which best allow us to capture the uncertainty of model estimation
in either case and permits a more natural comparison between approaches.

@@

@@

To illustrate the value of the non-parametric Bayesian approach to management,
we focus on example of a system containing such a tipping point whose dynamics
can still be described by a simple, one-dimensional parametric model.  
We will focus on a simple parametric model for a single species [derived
from fist principles by @Allen2005a] as our underlying "reality".
X_{t+1} &= Z_t f(S_t) \\
S_t &= X_t - h_t \\
f(S_t) &= S_t e^{r \left(1 - \frac{S_t}{K}\right)\left(S_t - C\right)}
\end{align}

@@

In our example, we focus on the problem in which a manager must set 
the harvest level for a marine fishery each year to maximize the net
present value of the resource, given an estimated stock size from the year
before. The economic value and ecological concern have made marine fisheries the crucible for much
of the founding work [@Gordon1954; @Reed1979; @May1979; @Ludwig1982]
in managing ecosystems under uncertainty.  Global trends [@Worm2006]
and controversy [@Hilborn2007; @Worm2009] have made understanding these
challenges all the more pressing.

@@

A crux of this approach is correctly specifying the functional form of $f$,
along with its parameters.  The standard approach uses one of a 
handful of common parametric models representing the stock-recruitment
relationship, usually after estimating the model parameters from any 
available existing data. Uncertainty in the parameter estimates can 
be estimated and integrated over to determine the optimal policy under
under uncertainty [@Mangel1985; @Schapaugh2013]. Uncertainty in the model
structure itself can only be addressed in this approach by hypothesizing 
alternative model structures, and then performing some model choice or
model averaging  [@Williams2001; @Athanassoglou2012].

### Parametric Models

We consider three candidate parametric models for the stock-recruitment function, 
which we refer to by the first authors of the publications in which they were
first proposed.

We generate the data with a four-parameter model that contains
a tipping point (an Allee effect, see [@Allen, @Courchamp]) below which the stock
decreases to zero,

$$X_{t+1} = Z_t X_t e^{r \left(1 - \frac{S_t}{K}\right)\left(\frac{S_t- \theta}{K}\right)} \label{Allen} $$

$$ S_t = X_t - h_t $$

The parameter $C$ reflects the location of the tipping point, $K$ the carrying capacity of
the stock, and $r$ the base recruitment rate.  $S_t$ represents the stock size after
a harvest $h_t$ has been implemented.  $Z_t$ represents a log-normal random variable of
log-mean zero and log-standard deviation parameter $\sigma$.

We consider two alternative candidate models: the Ricker [@Ricker] stock-recruitment curve,

$$X_{t+1} = Z_t X_t e^{r \left(1 - \frac{S_t}{K} \right) } \label{Ricker} $$

and an alternative four-parameter model adapted from @Myers,

$$ X_{t+1} = Z_t \frac{r S_t^{\theta}}{1 - \frac{S_t^\theta}{K}} \label{Myers} $$

which contains a tipping point for $\theta > 2$ and becomes a Beverton-Holt model 
at $\theta = 1$.

<!-- Note that while similarly-named parameters contain similar connotations, the 
estimate and precise intepretation will be different in each model. -->

### Bayesian Inference of Parametric models

Given the sample data, we infer posterior distributions for each of the three
models listed above using a Markov Chain Monte Carlo Gibbs Sampler (jags, see 
appendix for implementation details and code) given uniform priors.  We run six
chains for $10^6$ steps each and then assess convergence by Gelman-Rubin criterion
and inspection of the traces, see appendix.

<!-- By transforming variables we can integrate out the the $\sigma$ parameter
and the constant term in the log-mean, see appendix.  The resulting one or
two parameters can then be evaluated over a vector or grid of possible values,
and the result transformed back to determine the posterior density of the parameters,
This approach avoids any concerns of convergence in the MCMC analysis and can thus
be used to verify that the performance of the parametric models can not be explained
by convergence issues.  However, this approach does not generalize easily to richer models,
such as the introduction of measurement noise or additional parameters, in which 
MCMC based approaches are the obvious choice.  -->

@@

The use of Gaussian process (GP) regression (or "kriging" in the geospatial
literature) to formulate a predictive model is relatively new in the
context of modeling dynamical systems [@Kocijan2005] and introduced
in the ecological modeling and fisheries management by @Munch2005.
An accessible and thorough introduction to the formulation and use of
GPs can be found in @Rasmussen2006.
